Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-10-14 Nicolas.tex
MS comments

\newcommand{\nn}[1]{\textcolor[rgb]{0.25, 0.50, 0}{#1}}
\newcommand{\mr}[1]{\textcolor[rgb]{0.5, 0.20, 0.3}{#1}}
\newcommand{\prob}[2]{\mathcal{P}\left( #1 \mid #2\right)}

@@

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and J. Nordin\inst{4}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{4}
    \orcid{0000-0002-3321-1432}
}

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to a $z_{\lim}=0.24$. However,
the SDSS surveys were more sensitive to limited spectroscopic resources;
\cite[][see their Section~2]{kessler2009} pointed out that during the first year
of SDSS, SNe~Ia with $r<20.5$~mag were favored for spectroscopic follow-up,
corresponding to a redshift cut at $0.15$. For the rest of the SDSS survey,
additional spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{\lim}=0.2$. Following
these analyses, we will use $z_{\lim}=0.20$ as the baseline SDSS redshift limit,
\nn{and the first 0.15 cut to compare the results of our modeling on a more
restrictive sample.}

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits roughly correspond to the peak of these
histograms. \nn{We test in Section~\ref{ssec:verify} whether we can consider
this sample to be free of selection effects. The results exposed} in
Section~\ref{sec:results} further supports that this sample selection indeed
provides a subset of SNe~Ia with insignificant selection effects when
compared to state-of-the-art Malmquist correction techniques.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}, \nn{where the
disappearance of the low-stretch population (typically $x_1 < -1$) is what we
test in this paper}. The evolution of the mean stretch is also shown in
Fig.~\ref{fig:modelall}, where the data are split in redshift bins of regular
sample size. We see that SNe~Ia at higher redshift have on average larger
stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift ($-0.17\pm
0.10$ at $z\sim0.05$), suggesting that the underlying stretch distribution is
drifting.

@@

In this analysis, we have built a volume-limited sample from a set of magnitude
limited ones, using a simple redshift limit. \nn{Even though it has the strength
to be independent of potential inaccuracy in the modeling of the complex
selection functions of the aforementioned dataset, it is suboptimal.
That fact lead to the use of conservative cuts as first indicators that there is
a drift that can't be explained by selection effects alone and a}n ongoing
follow up analysis will make use of the SNANA package \citep{SNANA} to account
for more subtle selection functions; \nn{however}, it is still necessary to test
if the sample we have created could show sign of leftover sample selection.

To do so, we compare the stretch and color distributions of the SNe~Ia in our
sample between datasets of overlapping redshifts. If indeed we have built a
volume-limited sample, the distributions of any SNe~Ia parameter at a given
redshift should not depend of their survey or origin. When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible. The two samples that overlap the most in redshift are PS1 and SDSS
in the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus the
most affected by potential selection effect leftover (see the corresponding
discussion in section~\ref{sec:sample}). In that same redshift range, PS1 has 52
SNe~Ia, which are at in the lowest redshift bin and thus unlikely to have any
selection issue. The Fig.~\ref{fig:distrib} present the stretch and color
distribution of both, together with p-values of Kolmogorov-Smirnov tests that
\nn{probe whether} the $x_1$ and $c$ of the SDSS and PS1 samples at $0.10 < z <
0.20$ arise from the same underlying distribution. Those p-values are high
($>10\%$), and if one cannot conclude from this test that these SNe~Ia are from
the same underlying distribution, one has no ground to claim they are not. This
is in agreement with the visual impression from Fig.~\ref{fig:distrib}. A
similar analysis \nn{was done} for PS1 and SNLS in the redshift range $0.20 < z
< 0.31$ where the same conclusion \nn{can} be drawn, i.e., there is no sign of
discrepancy in the stretch and color distributions between the low- and high-end
of the SNLS and PS1 datasets, respectively. Nonetheless, the small size of the
SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1) limits the interest of the
test as only strong deviation would be noticeable. Extending the redshift range
to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3) enables to get 31
SNe~Ia from SNLS in total, the stretch and color p-values remains high, at 34\%
and 4\% respectively.

@@

In this paper, we considered a simple Gaussian mixture modeling, but additional
data free from significant Malmquist bias would enable us to refine it as
necessary. We note that samples at the low- and high-redshift ends of the Hubble
diagram would be particularly helpful for this drifting analysis; fortunately
this will soon be provided by the Zwicky Transient Facility
\citep[low-$z$,][]{bellm2019, graham2019}, and Subaru and SeeChange SNe~Ia
programs (high-$z$), respectively. \nn{Moreover, a more detailed analysis using
the SNANA framework is currently in progress in order to account for more subtle
selection functions and refine the dataset.}

>>>>

%\newcommand{\nn}[1]{\textcolor[rgb]{0.25, 0.50, 0}{#1}}
\newcommand{\nn}[1]{\textcolor[rgb]{0.4, 0.8, 0}{#1}}
\newcommand{\mr}[1]{\textcolor[rgb]{0.5, 0.20, 0.3}{#1}}
\newcommand{\prob}[2]{\mathcal{P}\left( #1 \mid #2\right)}

@@

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and J. Nordin\inst{4}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{2,5}
    \orcid{0000-0002-3321-1432}
}

@@

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to a $z_{\lim}=0.24$. However,
the SDSS surveys were more sensitive to limited spectroscopic resources;
\cite[][see their Section~2]{kessler2009} pointed out that during the first year
of SDSS, SNe~Ia with $r<20.5$~mag were favored for spectroscopic follow-up,
corresponding to a redshift cut at $0.15$. For the rest of the SDSS survey,
additional spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{\lim}=0.2$. Following
these analyses, we will use $z_{\lim}=0.20$ as the baseline SDSS redshift limit,
\nn{and $z_{\lim}=0.15$ to compare the results of our modeling on a more
restrictive sample.}

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits roughly correspond to the peak of these
histograms. \nn{We test in Section~\ref{ssec:verify} whether we can consider
this sample to be free of selection effects. The consistency of the results for the two samples, discussed} in
Section~\ref{sec:results}, further supports that this sample selection indeed
provides a subset of SNe~Ia with insignificant selection effects when
compared to state-of-the-art Malmquist correction techniques.

@@

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \nn{Proportionally few low-stretch SNe (typically $x_1 < -1$) are observed at $z>0.4$}. The evolution of the mean stretch is also shown in
Fig.~\ref{fig:modelall}, where the data are split in redshift bins of regular
sample size. We see that SNe~Ia at higher redshift have on average larger
stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift ($-0.17\pm
0.10$ at $z\sim0.05$), suggesting that the underlying stretch distribution is
drifting.

@@

In this analysis, we have built a volume-limited sample from a set of magnitude
limited ones, using a simple redshift limit. \nn{The advantage of this approach is that our results are robust to any potential inaccuracy in the modeling of the complex
selection functions for the individual datasets. However, while we will find consistent results when considering a sample derived from conservative cuts (Section~\ref{sec:results}), there remains the possibility that our sample is not fully volume-limited and that unaccounted for selection effects remain. To fully account for this possibility, a} follow up analysis will make use of the SNANA package \citep{SNANA} to account
for more subtle selection functions; \nn{however}, it is still necessary to test
if the sample we have created could show sign of leftover sample selection.
%While our conservative comparision sample is That fact lead to the use of conservative cuts as first indicators that there isa drift that can't be explained by selection effects alone and a}n ongoing

To do so, we compare the stretch and color distributions of the SNe~Ia in our
sample between datasets of overlapping redshifts. If indeed we have built a
volume-limited sample, the distributions of any SNe~Ia parameter at a given
redshift should not depend of their survey or origin. When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible. The two samples that overlap the most in redshift are PS1 and SDSS
in the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus the
most affected by potential selection effect leftover (see the corresponding
discussion in section~\ref{sec:sample}). In that same redshift range, PS1 has 52
SNe~Ia, which are at in the lowest redshift bin and thus unlikely to have any
selection issue. The Fig.~\ref{fig:distrib} present the stretch and color
distribution of both, together with p-values of Kolmogorov-Smirnov tests that
\nn{probe whether} the $x_1$ and $c$ of the SDSS and PS1 samples at $0.10 < z <
0.20$ arise from the same underlying distribution. Those p-values are high
($>10\%$), and if one cannot conclude from this test that these SNe~Ia are from
the same underlying distribution, one has no ground to claim they are not. This
is in agreement with the visual impression from Fig.~\ref{fig:distrib}. \nn{We perform a}
similar analysis for PS1 and SNLS in the redshift range $0.20 < z
< 0.31$ where the same conclusion \nn{can} be drawn, i.e., there is no sign of
discrepancy in the stretch and color distributions between the low- and high-end
of the SNLS and PS1 datasets, respectively. Nonetheless, the small size of the
SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1) limits the interest of the
test as only strong deviation would be noticeable. Extending the redshift range
to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3) enables to get 31
SNe~Ia from SNLS in total, the stretch and color p-values remains high, at 34\%
and 4\% respectively.

@@

In this paper, we considered a simple Gaussian mixture modeling, but additional
data free from significant Malmquist bias would enable us to refine it as
necessary. We note that samples at the low- and high-redshift ends of the Hubble
diagram would be particularly helpful for this drifting analysis; fortunately
this will soon be provided by the Zwicky Transient Facility
\citep[low-$z$,][]{bellm2019, graham2019}, and Subaru and SeeChange SNe~Ia
programs (high-$z$), respectively. \nn{To fully break the degeneracy between selection effects and population evolution, an analysis using
the SNANA framework is underway. This analysis will allow for subtle selection effects to be considered, and by removing the need for a redshift cut, increase the redshift leverage and statistical significance of our results.}
